AusTraits, a curated plant trait database for the Australian flora

### 2021-07-18 ms.Rmd
Revise paper
- updated reference list
- reformat references (don't cite unpublished, correct some citations)
- change online table to include dataset ids
- more accurate description of differences to TRY

```{r, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
devtools::load_all()
load_data()

@@

\normalsize

@@

Species traits are essential for comparing ecological strategies among plants, both within any given vegetation, and also arrayed across environmental space and evolutionary lineages [@diaz2016global; @kunstler2016plant; @zanne2014three; @cornwell2014functional]. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function [@chapin1993evolution; @adler2014functional; @diaz1998plant,violle2007let]. Traits thereby provide useful indicators of species' behaviours in communities and ecosystems, regardless of their taxonomy [@Westoby-1998; @violle2007let; @funk2017revisiting]. Through global initiatives the volume of available trait information for plants has grown rapidly in the last two decades [@Kattge2011try; @Kattge-2020]. However, the geographic coverage of trait measurements across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions, and, more generally, development of theory accounting for the diversity of plant strategies.

@@

Here we introduce the AusTraits database (hereafter AusTraits), a compilation of plant traits for the Australian flora. Currently, AusTraits draws together `r austraits$sources %>% length()` primary sources and contains `r austraits$traits %>% nrow()` measurements spread across `r austraits$traits$trait_name %>% n_distinct()` different traits for  `r austraits$traits$taxon_name %>% n_distinct()` taxa. To assemble AusTraits from diverse primary sources and make data available for reuse, we needed to overcome three main types of challenges (Figure 1): 1) Accessing data from diverse original sources, including field studies, online databases, scientific articles, and published taxonomic floras; 2) Harmonising these diverse sources into a federated resource, with common taxon names, units, trait names, and data formats; and 3) Distributing versions of the data under suitable license. To meet this challenge, we developed a workflow which draws on emerging community standards and our collective experience building trait databases.

@@

AusTraits has been developed as a standalone database, rather than as part of the existing global database TRY [@Kattge-2020], for three reasons. First, we sought to establish an engaged and localised community, actively collaborating to enhance coverage of plant trait data within Australia. It was our opinion that a community would form more readily to fill gaps in national knowledge of traits with local ownership of the resource. While we will never have a counterfactual, a vibrant community excited to be part of this initiative has indeed been established and coverage is much higher for Australian species than has been achieved since TRY’s inception. Local ownership also aligns well with funding opportunities and national research priorities, and enables database coordinators to progress at their own speed. Second, we wanted to implement new features not available in TRY. For instance, AusTraits is fully reproducible, meaning the entire database can be rebuilt from its original sources at any point. This has enabled us to actively curate the dataset to reduce duplications and update mapping of traits as contributors provide feedback and trait definitions and values are edited. Moreover, the curation is achieved on GitHub, such that the history of all changes is traceable and, where necessary, reversible. Third, AusTraits will be distributed via different mechanisms to TRY. Data from TRY is requested via a standalone portal. By contrast, versions of AusTraits are being regularly archived on a third party data repository (Zenodo), built for long term data archiving and an established API. This will allow for active connection to a variety of end users including, but not only, TRY.

We envision AusTraits as an on-going collaborative initiative for easily archiving and sharing trait data about the Australian flora. Open access to a comprehensive resource like this will generate significant new knowledge about the Australian flora across multiple scales of interest, as well as reduce duplication of effort in the compilation of plant trait data, particularly for research students and government agencies seeking to access information on traits. In coming years, AusTraits will continue to be expanded, with integrations into other biodiversity platforms and expansion of coverage into historically neglected plant lineages in trait science, such as pteridophytes (lycophytes and ferns). Further, through international initiatives, such as the Open Traits Network and TraitNet, linkages are being forged between plant datasets and a variety of other organismal databases [@Naeem-2009; @Gallagher-2020].

@@

AusTraits version `r austraits$build_info$version` was assembled from `r austraits$sources %>% length()` distinct sources, including published papers, field measurements, glasshouse and field experiments, botanical collections, and taxonomic treatments. Initially we identified a list of candidate traits of interest, then identified primary sources containing measurements for these traits, before contacting authors for access. As the compilation grew, we expanded the list of traits considered to include any measurable quantity that had been quantified for at least a moderate number of taxa (n > 20).

@@

A full list of traits and their sources appears in Supplementary Table 1 `r n_records$refs %>% str_split(", ") %>% unlist() %>% sort() %>% unique %>% paste(collapse = "; @") %>% paste0("[@", ., "]")`. This list was developed gradually as new datasets were incorporated, drawing from original source publications and a published thesaurus of plant characteristics [@Garnier-2017]. We categorised traits based on the tissue where it is measured (`r trait_categories$tissue %>% unique() %>% sort() %>% paste(collapse=", ")`) and the type of measurement (`r trait_categories$category %>% unique() %>% sort() %>% gsub("_", " ", .) %>% paste(collapse=", ")`). Version `r austraits$build_info$version` of AusTraits includes  `r n_records$Type %>% table() %>% sort(decreasing = T) %>% paste(.,  names(.), collapse= " and ")` traits.

@@

An R package is being developed (see [github.com/traitecoevo/austraits](http://github.com/traitecoevo/austraits)) to provide easy access and guidance on usage for those using this platform.

@@

```{r}
#As of May 2020, the APC included 28,981 accepted species, subspecies, varieties and forma.  
```

@@

The most common traits are non geo-referenced records from floras; these are trait values representing a continental or region mean (or spread) and hence are not linked to a location. Yet, geo-referenced records were available in several traits for more than 10% of the flora (Figure 3a). Coverage is notably higher for geo-referenced measurements of some tissues and trait types - such as bark stems and roots - relative to non-georeferenced measurements (Figure 3).

@@

We implemented three strategies to maintain data quality. First, we conducted a detailed review of each source based on a bespoke report, showing all data and metadata, by both an AusTraits curator and the original contributor (where possible). Measurements for each trait were plotted against all other values for the trait in AusTraits, allowing quick identification of outliers. Corrections suggested by contributors were combined back into AusTraits and made available with the next release. Version `r austraits$build_info$version` of AusTraits, described here, is the sixth release.

Second, we implemented automated tests for each dataset, to confirm that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of accepted values maintained by the curators. Data that did not pass these tests were moved to a separate spreadsheet (“excluded_data”) that is also made available for use and review.

Third, we provide a pathway for user feedback. AusTraits is a community resource and we encourage engagement from users on maintaining the quality and usability of the dataset. As such, we welcome reporting of possible errors, as well as additions and edits to the online documentation for AusTraits that make using the existing data, or adding new data, easier for the community. Feedback can be posted as an issue directly at the projects GitHub page (http://traitecoevo.github.io/austraits.build).

@@

Using the taxon names aligned with the APC, data can be queried against location data from the Atlas of Living Australia. To create the phylogenetic tree in Figure 6, we pruned a master tree for all higher plants [@Smith-2018] using the package `V.PhyloMaker` [@Jin-2020] and visualising via `ggtree` [@Yu-2017]. To create Figure 3A, we used the package `plotbiomes` [@Stefan-2020] to create the baseline plot of biomes.

@@

All code, raw and compiled data are hosted within GitHub repositories under the Trait Ecology and Evolution organisation (http://traitecoevo.github.io/austraits.build/). The archived material includes all data sources and code for rebuilding the compiled dataset. The code used to produce this paper is available at http://github.com/traitecoevo/austraits_ms. (All code will be made available prior to final publication.)

@@

We acknowledge the work of all Australian taxonomists and their supporting institutions, whose long-term work on describing the flora has provided a rich source of data for AusTraits, including: Australian National Botanic Gardens; Australian National Herbarium; Biodiversity Science, Parks Australia; Centre for Australian National Biodiversity Research; Department of Biodiversity, Conservation and Attractions, Western Australia; Department of Environment, Land, Water and Planning, Victoria; Flora of Australia; Kew; National Herbarium of NSW; National Herbarium of Victoria; Northern Territory Herbarium; NSW Department of Planning, Industry, and Environment; Queensland Herbarium; State Herbarium of South Australia; Tasmanian Herbarium; and the Western Australian Herbarium. We gratefully acknowledge input from the following persons who contributed to data collection `r acknowledge$name %>% unique() %>% paste( collapse= ", ")`. We thank H Cornelissen, H Poorter, SC McColl-Gausden, and one anonymous reviewer for feedback on an earlier draft, and K Levett for advice on data structures. This work was supported by fellowship grants from Australian Research Council to Falster (FT160100113), Gallagher (DE170100208) and Wright (FT100100910), a grant from Macquarie University to Gallagher, and grants from the Australian Research Data Commons (ARDC), via their "Transformative data collections" ([doi:10.47486/TD044](https://doi.org/10.47486/TD044)) and "Data Partnerships" ([doi: 10.47486/DP720](https://doi.org/10.47486/DP720)) programs. The ARDC is enabled by National Collaborative Research Investment Strategy (NCRIS).

@@

\clearpage

@@

```{r}
# Supplementary Table 1
 n_records %>%
  mutate(
         Trait = txt_wrap(Trait),
         Group = paste0(str_to_sentence(Tissue), " (", Category, ")"), 
         Type = ifelse(Type=="numeric", "num.", Type),
         Type = ifelse(Type=="categorical", "cat.", Type),
         Type = ifelse(Type=="character", "chr.", Type),
         refs = purrr::map_chr(refs, format_cites_n),
         ) %>%
  select(-Tissue, -Category) %>%
  select(Group, everything()) %>%
  split(., .$Group) %>% purrr::map(~bind_rows(tibble(Group=.x$Group[1]), .x)) %>% bind_rows() %>% mutate_all(~replace_na(.x, "")) %>%
  mutate(Group = ifelse(duplicated(Group), "", Group)) ->x

@@

  x %>% select(Group, Trait, refs) %>% 
    mutate_all(~paste(.x, ";") %>% gsub("\\\\newline ", "",.)) %>%
    knitr::kable()
```

\clearpage

\footnotesize

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
devtools::load_all()
load_data(update=TRUE)

@@

@@

Species traits are essential for comparing ecological strategies among plants, both within any given vegetation and across environmental space or evolutionary lineages [@zanne2014three; @cornwell2014functional; @diaz2016global; @kunstler2016plant]. Broadly, a trait is any measurable property of a plant capturing aspects of its structure or function [@chapin1993evolution; @adler2014functional; @diaz1998plant, @violle2007let]. Traits thereby provide useful indicators of species' behaviours in communities and ecosystems, regardless of their taxonomy [@Westoby-1998; @violle2007let; @funk2017revisiting]. Through global initiatives the volume of available trait information for plants has grown rapidly in the last two decades [@Kattge2011try; @Kattge-2020]. However, the geographic coverage of trait measurements across the globe is patchy, limiting detailed analyses of trait variation and diversity in some regions, and, more generally, development of theory accounting for the diversity of plant strategies.

@@

Here we introduce the AusTraits database (hereafter AusTraits), a compilation of plant traits for the Australian flora. Currently, AusTraits draws together `r austraits$traits$dataset_id %>% n_distinct()` distinct sources and contains `r austraits$traits %>% nrow()` measurements spread across `r austraits$traits$trait_name %>% n_distinct()` different traits for  `r austraits$traits$taxon_name %>% n_distinct()` taxa. To assemble AusTraits from diverse primary sources and make data available for reuse, we needed to overcome three main types of challenges (Figure 1): 1) Accessing data from diverse original sources, including field studies, online databases, scientific articles, and published taxonomic floras; 2) Harmonising these diverse sources into a federated resource, with common taxon names, units, trait names, and data formats; and 3) Distributing versions of the data under suitable license. To meet this challenge, we developed a workflow which draws on emerging community standards and our collective experience building trait databases.

@@

AusTraits has been developed as a standalone database, rather than as part of the existing global database TRY [@Kattge-2020], for three reasons. First, we sought to establish an engaged and localised community, actively collaborating to enhance coverage of plant trait data within Australia. We envisioned  that a community would form more readily to fill gaps in national knowledge of traits with local ownership of the resource. While we will never have a counterfactual, a vibrant community excited to be part of this initiative has indeed been established and coverage is much higher for Australian species than has been achieved since TRY’s inception. Local ownership also aligns well with funding opportunities and national research priorities, and enables database coordinators to progress at their own speed. Second, we wanted to apply an entirely open-source approach to the aggregation workflow [following @Falster-2015]. All the code and raw files used to create the compiled database are available, and this database is freely available via a third party data repository (Zenodo) which is itself built for long term data archiving, with an established API. Finally, we targeted primary data sources, where possible; whereas TRY accepts aggregated datasets. The hope was that this would increase data quality, by removing intermediaries and easier identification of duplicates.

While independent, the overall structure of AusTraits is similar to that of TRY, ensuring the two databases will be interoperable. Both databases are founded on similar principles and terminology [@madin2007ontology; @Garnier-2017]. Increasingly, researchers and biodiversity portals are seeking to connect diverse datasets [@Gallagher-2020], which is possible if they share a common foundation.

We envision AusTraits as an on-going collaborative initiative for easily archiving and sharing trait data about the Australian flora. Open access to a comprehensive resource like this will generate significant new knowledge about the Australian flora across multiple scales of interest, as well as reduce duplication of effort in the compilation of plant trait data, particularly for research students and government agencies seeking to access information on traits. In coming years, AusTraits will continue to be expanded, with integrations into other biodiversity platforms and expansion of coverage into historically neglected plant lineages in trait science, such as pteridophytes (lycophytes and ferns). Further, through international initiatives, such as the Open Traits Network, linkages are being forged between plant datasets and a variety of other organismal databases [@Gallagher-2020].

@@

AusTraits version `r austraits$build_info$version` was assembled from `r austraits$traits$dataset_id %>% n_distinct()` distinct sources, including published papers, field measurements, glasshouse and field experiments, botanical collections, and taxonomic treatments. Initially we identified a list of candidate traits of interest, then identified primary sources containing measurements for these traits, before contacting authors for access. As the compilation grew, we expanded the list of traits considered to include any measurable quantity that had been quantified for at least a moderate number of taxa (n > 20).

@@

A full list of traits and their sources appears in Supplementary Table 1 `r c(n_records$refs_primary_pub, n_records$refs_secondary_pub) %>% collapse("; @") %>% paste0("[@", ., "]")` `r c(n_records$unpub_people) %>% collapse(", ") %>% paste0("(", ., "; unpublished data)")`. This list was developed gradually as new datasets were incorporated, drawing from original source publications and a published thesaurus of plant characteristics [@Garnier-2017]. We categorised traits based on the tissue where it is measured (`r trait_categories$tissue %>% unique() %>% sort() %>% paste(collapse=", ")`) and the type of measurement (`r trait_categories$category %>% unique() %>% sort() %>% gsub("_", " ", .) %>% paste(collapse=", ")`). Version `r austraits$build_info$version` of AusTraits includes  `r n_records$Type %>% table() %>% sort(decreasing = T) %>% paste(.,  names(.), collapse= " and ")` traits.

@@

The R package `austraits` (https://github.com/traitecoevo/austraits) provides easy access to data and examples on manipulating data (e.g. joining tables, subsetting) for those using this platform.

@@

@@

The most common traits are non geo-referenced records from floras; these are trait values representing a continental or region mean (or spread) and hence are not linked to a location. Yet, geo-referenced records were available in several traits for more than 10% of the flora (Figure 3a). Coverage is notably higher for geo-referenced measurements of some tissues and trait types - such as bark stems and roots - relative to non-geo-referenced measurements (Figure 3).

@@

We implemented three strategies to maintain data quality. First, we conducted a detailed review of each source based on a bespoke report, showing all data and metadata, by both an AusTraits curator (primarily Wenk) and the original contributor (where possible). Measurements for each trait were plotted against all other values for the trait in AusTraits, allowing quick identification of outliers. Corrections suggested by contributors were combined back into AusTraits and made available with the next release. Version `r austraits$build_info$version` of AusTraits, described here, is the sixth release.

Second, we implemented automated tests for each dataset, to confirm that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of allowed values. Data that did not pass these tests were moved to a separate spreadsheet (“excluded_data”) that is also made available for use and review.

Third, we provide a pathway for user feedback. AusTraits is an open-source community resource and we encourage engagement from users on maintaining the quality and usability of the dataset. As such, we welcome reporting of possible errors, as well as additions and edits to the online documentation for AusTraits that make using the existing data, or adding new data, easier for the community. Feedback can be posted as an issue directly at the project's GitHub page (http://traitecoevo.github.io/austraits.build).

@@

Using the taxon names aligned with the APC, data can be queried against location data from the Atlas of Living Australia. To create the phylogenetic tree in Figure 6, we pruned a master tree for all higher plants [@Smith-2018] using the package `V.PhyloMaker` [@Jin-2020] and visualising via `ggtree` [@Yu-2017]. To create Figure 3a, we used the package `plotbiomes` [@Stefan-2020] to create the baseline plot of biomes.

@@

All code, raw and compiled data are hosted within GitHub repositories under the Trait Ecology and Evolution organisation (http://traitecoevo.github.io/austraits.build/). The archived material includes all data sources and code for rebuilding the compiled dataset. The code used to produce this paper is available at http://github.com/traitecoevo/austraits_ms.

@@

We acknowledge the work of all Australian taxonomists and their supporting institutions, whose long-term work on describing the flora has provided a rich source of data for AusTraits, including: Australian National Botanic Gardens; Australian National Herbarium; Biodiversity Science, Parks Australia; Centre for Australian National Biodiversity Research; Department of Biodiversity, Conservation and Attractions, Western Australia; Department of Environment, Land, Water and Planning, Victoria; Flora of Australia; Kew; National Herbarium of NSW; National Herbarium of Victoria; Northern Territory Herbarium; NSW Department of Planning, Industry, and Environment; Queensland Herbarium; State Herbarium of South Australia; Tasmanian Herbarium; and the Western Australian Herbarium. We gratefully acknowledge input from the following persons who contributed to data collection `r acknowledge$name %>% unique() %>% paste( collapse= ", ")`. We thank H Cornelissen, H Poorter, SC McColl-Gausden, and one anonymous reviewer for feedback on an earlier draft, and K Levett for advice on data structures. This work was supported by fellowship grants from Australian Research Council to Falster (FT160100113), Gallagher (DE170100208) and Wright (FT100100910), a grant from Macquarie University to Gallagher, and grants from the Australian Research Data Commons (ARDC), via their "Transformative data collections" (https://doi.org/10.47486/TD044) and "Data Partnerships" (https://doi.org/10.47486/DP720) programs. The ARDC is enabled by National Collaborative Research Investment Strategy (NCRIS).

@@

@@

```{r}
format_cites <- function(x) {
  # Make sorted vector of all unique refs
  str_split(x, ", ") %>% unlist() %>% sort() %>% 
    paste(collapse = "; @") %>% paste0("[ [@", ., "] ]" )
}

# Supplementary Table 1
 n_records %>%
  mutate(
         Trait = txt_wrap(Trait),
         Group = paste0(str_to_sentence(Tissue), " (", Category, ")"), 
         Type = ifelse(Type=="numeric", "num.", Type),
         Type = ifelse(Type=="categorical", "cat.", Type),
         Type = ifelse(Type=="character", "chr.", Type),
         Sources_primary = ifelse(refs_primary_pub =="","",
                         purrr::map_chr(refs_primary_pub, format_cites)),
         Sources_secondary = ifelse(refs_secondary_pub =="","",
                         purrr::map_chr(refs_secondary_pub, format_cites))
         ) %>%
  select(-Tissue, -Category, -starts_with("refs"), -unpub_people) %>%
  select(Group, everything()) %>%
  split(., .$Group) %>% purrr::map(~bind_rows(tibble(Group=.x$Group[1]), .x)) %>% bind_rows() %>% mutate_all(~replace_na(.x, "")) %>%
  mutate(Group = ifelse(duplicated(Group), "", Group)) ->x

@@

  x %>% select(Group, Trait, Sources_primary, Sources_secondary) %>% mutate(Newline="XXX") %>% 
    mutate_all(~paste(.x, ";") %>% gsub("\\\\newline ", "",.)) %>%
    knitr::kable()
```

## References
